---
title: "Magic the Gathering"
subtitle: "An analysis of the \"Modern\" metagame in April 2022"
author: "Massimo del Prato"
date: '27-06-2022'
output:
  ioslides_presentation: 
    widescreen: true
    smaller: true
  
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = T,warning = F, message = F, echo = FALSE}
library(cowplot)
library(magick)
library(knitr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(purrr)
library(magrittr)
library(igraph)
library(tidygraph)
library(ggraph)
library(ggrepel)
library(tidyverse)
library(RColorBrewer)
```

## Magic: The Gathering

Magic: The Gathering is a tabletop and digital collectible card game created by Richard Garfield released in 1993 by Wizards of the Coast. Currently is one of the most famous trading card game in the world.

```{r, echo=FALSE, fig.align="center"}
pic1 = ggdraw() + draw_image("./images/back.jpeg", scale = 1)
pic1
```

## The "Modern" format

Modern is one of the most popular format in the competitive setting of Magic: The Gathering.

It allows all cards printed from July 29, 2003 to the present to be played (with some exceptions).

A player must use a deck consisting of at least 60 cards plus 15 cards in the sideboard and it can use at most 4 copies of the same card, except for Basic Land cards that can be played in any number.

The analysis was carried out on all decks that ranked well in competitive tournaments with at least 8 players in the month of April 2022


## The dataset

The dataset that I have analyzed is one made by myself.
I have downloaded 798 deck lists from [mtgtop8](https://www.mtgtop8.com/format?f=MO&meta=236&a=) in txt format and I have arranged them in order to obtain a dataset where each entry has the following variable values:

* The **deck_name**
* The **player**
* The **deck_list**
* The **type**


## The dataset

The variable **deck_list** is a nested dataframe where an entry contains the information about 1159 different cards (got by [scryfall](https://scryfall.com/)):

* the **name** of the cards played in the deck
* the **number** of copies played
* the **colors** 
* and other data
```{r, echo=FALSE, fig.align="center"}
pic2 = ggdraw() + draw_image("./images/forestjacenivsphere.png", scale = 1)
pic2
```


## Objectives

* Find out the best players in the month of April and see if they are specialized in specific strategies or if they are versatile.

* Find out the most played cards and see how they relate with each other and with the deck played, and then the same for the less used cards and for signature cards.



```{r, include = T,warning = F, message = F, echo = FALSE}
# Load dataset 
load("complete_Dataset.Rdata")
df = my_dataframe
```


## Most played Deck type

From an analysis of the played decks, it can be seen that the most played deck type is the **AGGRO**.

```{r, include = T,warning = F, message = F, echo = FALSE}
all_decks = 
  df %>%
  select(deck_name, type)
all_decks = unique(all_decks)


# Toplogie di mazzi giocati
decks = 
  df %>%
  count(type) %>%
  rename(counts = n) %>%
  arrange(-counts)

ggplot(data = decks, mapping = aes(x = type, y = counts)) +
  geom_col(aes(fill = type)) +
  geom_text(aes(label = paste0(round(counts/798*100, digits = 2), "%")), vjust = -0.3) +
  theme_minimal()
```

## Top 10 used decks

```{r, include = T,warning = F, message = F, echo = FALSE}
# Top 10 played decks

top_decks = 
  my_dataframe %>%
  count(deck_name) %>%
  arrange(-n) %>%
  rename("played" = n)


top10 = head(top_decks, 10)


t = sum(top10$played)

df2 = top10 %>% 
  mutate(csum = rev(cumsum(rev(played))), 
         pos = played/2 + lead(csum, 1),
         pos = if_else(is.na(pos), played/2, pos))

ggplot(top10, aes(x = "" , y = played, fill = fct_inorder(deck_name))) +
  geom_col(width = 1, color = 1) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Paired") +
  geom_label_repel(data = df2,
                   aes(y = pos, label = paste0((round(played/t*100, digit = 2)), "%")),
                   size = 4.5, nudge_x = 1, show.legend = FALSE) +
  guides(fill = guide_legend(title = "Top 10 played Decks")) +
  theme_void()

```

## Top 10 used decks 

Despite the most popular deck type, the most popular deck is **4c Control** of the the type **CONTROL** which is the less popular deck type. 

Anyway more than the 40% of the 10 most played decks is an **AGGRO** deck.


##  Who are the best players?

The best players are the players that figure out more than once in the dataset. This is due to the fact that they have obtained a good result in more than a tournament in April. 

Apparently it was not easy to get more than a good result in April, as a matter of fact only 59 players out of 729 got more than one good result. And of these 59 only 10 got three.

```{r, include = T,warning = F, message = F, echo = FALSE, fig.height=4}
# Players and their number of results

players = 
  df %>%
  count(player) %>%
  arrange(-n) %>%
  rename(good_results = n)

ggplot(data = players, aes(x = good_results)) + 
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.3) +
  theme_minimal()
```



```{r, include = T,warning = F, message = F, echo = FALSE}
# The best players: players who appear twice in the dataset
best_players_2 = 
  df %>%
  count(player) %>%
  filter(n > 2) %>%
  arrange(-n)

best_players_1 = 
  df %>%
  count(player) %>%
  filter(n > 1) %>%
  arrange(-n)



# Deck played by the good players (>1)
best_player_decks_1 = 
  best_players_1 %>%
  inner_join(df, by = "player") %>%
  select(-n)

# Deck played by the best players (>2)
best_player_decks_2 = 
  best_players_2 %>%
  inner_join(df, by = "player") %>%
  select(-n)
  

#ggplot(data = best_player_decks_1) + 
#  geom_bar(mapping = aes(x = type, fill = deck_name)) + 
#  theme_classic()
```


## How specialized are the top players in terms of deck types?

In this section I have considered the 10 players with three results.

What I wanted to find out is whether to be a good player it was possible to play different types of decks or whether it was better to specialize on one strategy.

## How specialized are the top players in terms of deck types?

```{r, include = T,warning = F, message = F, echo = FALSE}
# Deck types played by the best players
bp_types = 
  best_player_decks_2 %>%
  select(-deck_list, -deck_name)

bp_types = bp_types[, c(2, 1)]

# Make a bipartite graph to show which type of deck do the best players play
mbp = table(bp_types)
ty_graph = graph.incidence(mbp, weighted = TRUE)
#is.bipartite(ty_graph)

#degree(ty_graph)

ty_lay = layout_as_bipartite(ty_graph)

set_graph_style()
ty_graph %>%
ggraph(layout = ty_lay) +
  geom_node_point(aes(color = type, size = 4, shape = type),  show.legend = FALSE) + 
  geom_edge_link(color = "grey") +
  geom_node_text(aes(label = name), repel = TRUE,  nudge_x = ty_graph$data$x * 1, nudge_y = ty_graph$data$y * .1) + 
  coord_flip()


```





## How specialized are the top players in terms of deck types?
Some interesting results:


* **Richard_Neal** has played two deck types.
* **Bosko_Milakovic** is the only one who managed to use all the three deck types. So I will say he's the most versatile player.
* All the other players are specialized on a single deck type.


So the tendency is to be specialized in a single deck type.


## How specialized are the top players in terms of decks played?

```{r, include = T,warning = F, message = F, echo = FALSE}
bp_decks = 
  best_player_decks_1 %>%
  select(-deck_list, -type) 
  

nodes = 
  bp_decks %>%
  group_by(player) %>%
  summarise(n_of_played_decks = n_distinct(deck_name))



p1 = bp_decks
p2 = p1
p3 = 
  inner_join(p1, p2, by = "deck_name") %>%
  filter(player.x != player.y)

# Remove the "symmetric" duplicate
p3 = p3[, c(1, 3, 2)]
p3[-3] = t(apply(p3[-3], 1, sort))
p3 = unique(p3)

edges = 
  p3 %>%
  rename(from = player.x, to = player.y) 

deck_graph = tbl_graph(nodes = nodes, 
                       edges = edges, 
                       directed = FALSE,
                       node_key = "player")

deck_lay = layout_nicely(deck_graph)

set_graph_style()
deck_graph %>%
  activate(nodes) %>%
  mutate(deg = centrality_degree()) %>%
ggraph(layout = deck_lay) + 
  geom_edge_link(aes(color = deck_name), show.legend = F) + 
  geom_node_point(aes(size = n_of_played_decks, color = deg)) + 
  geom_node_text(aes(label = ifelse(n_of_played_decks > 2 | deg > 18, player, NA)), repel = TRUE)


```


## How specialized are the top players in terms of decks?
Speaking about the single decks, I have analyzed all the player with more than one good result.

The relation must be interpreted as: x has played the same deck as y.

* There are only two players who have played three different decks: **Bosko_Milakovic** and **Ryan_Gassaway**
* **Bosko_Milakovic** is not only the most versatile player but he's also the most creative one. In fact he has played three decks that none of the other best players have played.
* **Jesse_James_Adams** is the player with the highest number of links as he has played the two most popular decks: **4c Control** and **UR Aggro**.

In term of different decks (that may be different but belong to the same type), the players are a bit less specialized with a good number of them which has played different deck.



```{r, include = T,warning = F, message = F, echo = FALSE}
# Move the focus on cards

load("./All mtg DataBase/allCardsDB.Rdata")

ac = allCards

card_pull =
  my_dataframe %>%
  unnest(deck_list)


onlyCards = 
  card_pull %>%
  select(-player, type) %>%
  filter(main_type != "Token Creature")

# All the single cards that are used in the analyzed decks 
# together with their information
sin = 
  onlyCards %>%
  count(name) %>%
  select(-n) %>%
  inner_join(ac, by = "name")



# The number fo different cards in the dataset
n_of_different_cards = nrow(sin)

# All the entry of the cards
total_list = 
  onlyCards %>%
  group_by(name) %>%
  arrange(name, by_group = TRUE) 


c_list = total_list %>%
  select(-deck_name, -use, -number)


# Cards ordered by the most played ones
most_used_cards =
  total_list %>%
  summarise(qnt = sum(number)) %>%
  arrange(-qnt) %>%
  inner_join(sin, by = "name") 


# Non-land cards ordered by the most played ones
most_nonLands = 
  most_used_cards %>%
  filter(main_type != "Land") %>%
  filter(main_type != "Basic Land") %>%
  filter(main_type != "Legendary Land") %>%
  filter(main_type != "Enchantment Land") %>%
  filter(main_type != "Basic Snow Land")
```


```{r, include = T,warning = F, message = F, echo = FALSE}
# Extracting the most 50 used cards
top_cards =
  head(most_nonLands, 50) %>%
  rename(color_identity = colors)
 

# Extracting the less used cards = cards which appear in a single copy in the dataset
worst_cards =
  most_nonLands %>%
  filter(qnt == 1) %>%
  rename(color_identity = colors)


top_cards$name = factor(top_cards$name,
                 levels = top_cards$name[order(top_cards$qnt, decreasing = FALSE)])


```


```{r, include = T,warning = F, message = F, echo = FALSE}
# Modify the color_identity variable for the top cards
tc = 
  top_cards %>%
  select(name, qnt, color_identity)

class(tc$color_identity) =  "character"

tc$color_identity = gsub("\"", "", as.character(tc$color_identity) )


tc$color_identity[tc$color_identity == "list()"] = "N"
tc$color_identity[tc$color_identity != "R" & 
                  tc$color_identity != "U" &
                  tc$color_identity != "G" &
                  tc$color_identity != "W" & 
                  tc$color_identity != "B" &
                  tc$color_identity !="N"] = "M"

```



## Which are the most played cards?
This graph shows the 20 most played non land cards. Quite surprisingly there are no black cards between them. However, it was predictable because the most played deck (**4c Control**) uses all colors except black. 

```{r, include = T,warning = F, message = F, echo = FALSE}

tc1 = head(tc, 20)

# Plot the 20 most used cards along with their quantity and color 
ggplot(tc1, aes(x = qnt, y = name, fill = color_identity, width = 0.85)) +
  geom_bar(color = "#343334", stat = "identity") +
  scale_fill_manual("card color", values = c("R" = "#f90000",
                                             "U" = "#1027ec",
                                             "G" = "#31c21d",
                                             "W" = "#ffffff", 
                                             "B" = "#000000", 
                                             "M" = "#ff9add",
                                             "N" = "#a79da4")) + 
  theme_minimal()


```

## Which are the most played cards?

To find the first black card we must look at the 25th position: **Thoughtseize** , anyway it is the only one black card in the top 40. 


```{r, include = T,warning = F, message = F, echo = FALSE}

tc2 = head(tc, 40)

# Plot the 40 most used cards along with their quantity and color 
ggplot(tc2, aes(x = qnt, y = name, fill = color_identity, width = 0.85)) +
  geom_bar(color = "#343334", stat = "identity") +
  scale_fill_manual("card color", values = c("R" = "#f90000",
                                             "U" = "#1027ec",
                                             "G" = "#31c21d",
                                             "W" = "#ffffff", 
                                             "B" = "#000000", 
                                             "M" = "#ff9add",
                                             "N" = "#a79da4")) + 
  theme_minimal()

```

## Correlation between the top 20 used cards

```{r, include = T,warning = F, message = F, echo = FALSE}
card_list = 
  total_list %>%
  select(deck_name, name)

# Selecting the top 10 popular cards
top20 = head(most_nonLands, 20)

# Get deck_name for the most popular cards
l = 
  inner_join(card_list, top20, by = "name") %>%
  select(deck_name, name)

# Card and deck names
l1 = unique(l)
l1 = l1[, c(2, 1)]

l2 =
  l1 %>%
  group_by(deck_name)
  
# Join to see in how many decks the top cards are used together
l3 = 
  inner_join(l1, l2, by = "deck_name") %>%
  filter(name.x != name.y) %>%
  rename(from = name.x, to = name.y) 
  

# Remove the "symmetric" duplicates 
l3 = l3[, c(1, 3, 2)]
l3[-3] = t(apply(l3[-3], 1, sort))
l3 = l3 %>%
  select(-deck_name)

l3 = unique(l3) 


# Vertices
vert =
  top20 %>%
  select(name)


# Create an undirected graph to study the correlation


card_graph = tbl_graph(nodes = vert,
                       edges = l3,
                       directed = FALSE,
                       node_key = "name"
                       )
 

card_lay = layout_nicely(card_graph) 

set_graph_style()
card_graph %>%
  activate(nodes) %>%
  mutate(degree = degree(card_graph)) %>%
ggraph(layout = card_lay) + 
  geom_edge_link(color = "grey") + 
  geom_node_point(aes(size = degree, alpha = degree)) + 
  geom_node_text(aes(label = name), repel = TRUE)
 

# Get the Degree Centrality to see which is the top cards which is more often use with other top cards
c = centr_degree(card_graph)
```


## Correlation between the top 20 used cards
How often the top cards are used together?

It results that almost all the top cards are used together at least in one the deck.  The only exception is **Shardless Agent** which has non links with 9 of the other top cards.

```{r, echo=FALSE, fig.align="center"}
pic2 = ggdraw() + draw_image("./images/shardless-agent.jpg", scale = 0.7)
pic2
```



## How decks exploit the top 20 cards

Apparently the use of the top cards is bind to the popularity of the deck.

```{r, include = T,warning = F, message = F, echo = FALSE}
td = 
  top_decks %>%
  inner_join(all_decks, by = "deck_name") %>%
  rename(deck_popularity = played, deck_type = type) %>%
  mutate(popularity_level = round(deck_popularity/798*100, digit = 2)) %>%
  arrange(-popularity_level) 


# Decks with the number of different cards played between the top 20 
top_in_decks = 
  l %>% 
  group_by(deck_name) %>%
  summarise(top_cards_played = n_distinct(name)) %>% 
  inner_join(td, by = "deck_name") 

ggplot(top_in_decks, aes(x = popularity_level, y = top_cards_played)) + 
  geom_point() + 
  geom_smooth() +
  theme_grey()

```





##  Which are the less played cards?

There are 142 cards that figure out in a single copy in all the analyzed decks.

```{r, include = T,warning = F, message = F, echo = FALSE, fig.align='center'}
w1 = 
  inner_join(card_list, worst_cards, by = "name") %>% 
  select(name, deck_name)

pic = ggdraw() + draw_image("./images/less.png", scale = 1)
pic
```

## Communities of less used cards


```{r, echo=FALSE, fig.align="center"}
# There are 142 singleton cards
n_worst_cards = nrow(w1)


# Copy to obtain the links
w2 = w1

# Join to see which singletons are played together
w3 = 
  inner_join(w1, w2, by = "deck_name") %>% 
  filter(name.x != name.y) %>%
  rename(from = name.x, to = name.y) 

w3 = w3[, c(1, 3, 2)]

# Remove the "symmetric" duplicates
w3[-3] = t(apply(w3[-3], 1, sort))
w3 = unique(w3)




# The graph of singleton cards
singles_graph = tbl_graph(nodes = w1,
                         edges = w3,
                         directed = FALSE, 
                         node_key = "name")

deg = degree(singles_graph)

singles_lay = layout_components(singles_graph)

set_graph_style()
singles_graph %>%
  ggraph(layout = singles_lay) +
  geom_edge_link(aes(color = deck_name)) + 
  geom_node_point() 

n_of_communities = clusters(singles_graph)$no
```




## Communities of less used cards

* There are 46 communities of single-copy-cards.
* 12 of them are composed of a single elements: they are the single-copy-cards which are played with no others single-copy-cards.
* **Enchantress** is the deck that uses the greatest number of single-copy-cards: 10.




## How decks exploits single-copy-cards
The tendency to play single-copy cards is more typical of less popular decks while top decks tend to maintain more standard lists.

The fact that a very popular deck plays single-copy-cards, in the case of **4c Control** which plays 5, is given by the fact that the deck is widely played and therefore it is easier to find different lists.

```{r, include = T,warning = F, message = F, echo = FALSE, fig.align='center', fig.height=4, fig.width=7}
# Number of single-copy-cards played by each deck
singles_per_deck = 
  w1 %>%
  group_by(deck_name) %>%
  count(deck_name) %>%
  inner_join(td, by = "deck_name") %>%
  rename(single_copies = n)

ggplot(singles_per_deck, aes(x = popularity_level, y = single_copies)) + 
  geom_point() + 
  theme_grey()

#View(singles_per_deck)
```




## The signature cards

The signature cards are cards that are played in a single deck but in any amount of copies. There are 579 signature cards but not all of them are interesting: in fact cards that are played in a deck which appear only once in the dataset are not so important.

For this analysis I have considered only the most important signature cards: cards that are played in decks that appear at least twice and are played in average at least 3 copies per deck. For a total of 83 signature cards.
```{r, echo=FALSE, fig.align="center"}
pic2 = ggdraw() + draw_image("./images/signatures.png", scale = 0.8)
pic2
```


```{r, include = T,warning = F, message = F, echo = FALSE}
qnt = 
  most_used_cards %>%
  select(name, qnt)


tl = 
  total_list %>% 
  select(name, deck_name)


# Cards that are played in a single deck  
tl1 =
  unique(tl) %>%
  count(name) %>%
  filter(n == 1)


tl2 = 
  inner_join(tl1, tl, by = "name")

tl2 = 
  unique(tl2) %>%
  select(-n) %>%
  inner_join(qnt, by = "name")

tl3 = 
  inner_join(tl2, td, by = "deck_name") %>%
  rename(copies_played = qnt) %>%
  mutate(average_copies_per_deck = round(copies_played / deck_popularity, digit = 2))




tl3 = unique(tl3)
tl3 = tl3[, c(1, 2, 5, 3, 4, 6, 7)]



s =
  tl3 %>%
  filter(deck_popularity > 2) %>%
  filter(average_copies_per_deck >= 3.00) %>%
  arrange(-average_copies_per_deck)
```

## Which decks use signature cards?

The results shows that the use of signature cards in not driven by the popularity of decks but by their type, in fact it seems that it is a feature more proper of the **COMBO** decks. 


```{r, include = T,warning = F, message = F, echo = FALSE}
most_important_signatures = 
  s %>%
  group_by(deck_name) %>%
  summarise(signature_cards_played = n_distinct(name)) %>%
  inner_join(td, by = "deck_name") 
#View(most_important_signatures)




ggplot(most_important_signatures, aes(x = popularity_level, y = signature_cards_played)) +
  geom_point(aes(color = deck_type, shape = deck_type)) + 
  theme_gray()
```



